---
title: "Black Rhinoceros Behaviour Dung Presentation Experiment"
author: "Ona Corberó Forn"
date: "2026-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lme4)
library(car)
library(emmeans)
```
# Behaviour Analysis of the dung experiment

First let's read the file and set up time in seconds properly

```{r}

beh <- read.csv2(paste0(wd$data, "behav_rhinos.csv"))
beh <- beh %>%  mutate(Time_sec = period_to_seconds(hms(Time))) # will say warning because not all lines have the time stamp, but it's okay actually 

```

## Sniffing frequency

Let's visualise how much sniffing frequency varies between the sex of the dung deposited, the sex of the rhino smelling it, and the paring 


```{r}

str(beh) # need to factorize

beh$Subject <- as.factor(beh$Subject)
beh$Subject_sex <- as.factor(beh$Subject_sex )
beh$Presentation_day <- as.factor(beh$Presentation_day)
beh$Pairing_Day <- as.factor(beh$Pairing_Day)
beh$Code <- as.factor(beh$Code)
beh$Event <- as.factor(beh$Event)
beh$Dungfrom <- as.factor(beh$Dungfrom )
beh$Dung_sex <- as.factor(beh$Dung_sex   )
beh$Pairing_sex  <- as.factor(beh$Pairing_sex )


#Need to know the sniff count per subject etc 
sniff_freq <- beh %>%
  filter(Code == "Sniff") %>%
  group_by(Subject, Dungfrom, Pairing_Day, Dung_sex, Pairing_sex, Subject_sex) %>%   
  summarise(Sniff_Count = n(), .groups = "drop")

fit <- glmer(  Sniff_Count ~ Pairing_sex + Pairing_Day + (1|Subject),
  family = poisson, data = sniff_freq)
Anova(fit) #not significant

fit <- glmer(  Sniff_Count ~ Dungfrom + (1|Subject),
  family = poisson, data = sniff_freq)
Anova(fit) #significant!


```
Some graphs for SNIFFING FREQUENCY

```{r}

#Dung sex 
a <-ggplot(sniff_freq, aes(x = Subject_sex, y = Sniff_Count, fill = Subject_sex)) +
    geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#a51c1c", "#f7ab47")) +
  labs(    x = "Subject Sex",
    y = "Sniffing Frequency" ) +
  theme_classic()  +  theme(legend.position = "none") + theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9)) 

#Pairing sex
z <-ggplot(sniff_freq, aes(x = Pairing_sex, y = Sniff_Count, fill = Pairing_sex)) +
   geom_boxplot( alpha = 0.7) +
  scale_fill_manual(values = c("#F0CC4A","#a7e940", "#2EB89A", "#6a6542")) +
  labs(     x = "Pairing Sex",
    y = "Sniffing Frequency"
  ) +   theme_classic()  +  theme(legend.position = "none") + theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9))

#Individual variation
j <-ggplot(sniff_freq, aes(y = Subject, x = Sniff_Count, fill = Subject)) +
    geom_boxplot(width = 0.5, alpha = 0.6) +
  scale_fill_manual(values = my_colors) +
  labs(       y = "Rhino ",
    x = "Sniff Frequency") +
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) + theme(legend.position = "none") + xlim(0, 17)


```


#Sniffing duration

Let's visualise how much sniffing time varies between the sex of the dung deposited, the sex of the rhino smelling it, and the sex paring 

```{r}

hist(beh$Time_sec) #lots of small numbers so poisson distribution 

sniff <- subset(beh, Code == "Sniff")
sniff$Code <- droplevels(sniff$Code)

fit <- glmer(Time_sec~ Dungfrom  +  (1 | Subject), family = poisson,  data = sniff)
Anova(fit) #very significant!

fit <- glmer(Time_sec~ Pairing_sex + Pairing_Day  + (1 | Subject), family = poisson,  data = sniff)
Anova(fit) # pairing sex significant as well as pairing day 

#post-hoc for pairing day 
emm_day <- emmeans(fit, ~ Pairing_Day)
pairs(emm_day, adjust = "tukey")


#post-hoc comparison for pairing sex
emm <- emmeans(fit, ~ Pairing_sex)
pairs(emm, adjust = "tukey") # so apparently there is no group that is significantly different from the others

plot(emm)

boxplot(Time_sec~ Pairing_sex , data = sniff)
#males spend more time smelling dung coming from males, but NOT significant (individual variation!)

boxplot(Time_sec~Subject, data = sniff) #BIG individual variation

#to know individual variation (intercept)
summary(fit)$varcor 


```

Some graphs for SNIFF DURATION

```{r}

# Dung sex and subject sex 

b <- ggplot(sniff, aes(x = Subject_sex, y = Time_sec, fill = Subject_sex)) +
   geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#a51c1c", "#f7ab47")) +
  labs(    x = "Subject Sex",
    y = "Sniffing Duration (s)") +
  theme_classic() +  theme(legend.position = "none") + theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9))


#Sex pairing 

y <- ggplot(sniff, aes(x = Pairing_sex, y = Time_sec, fill = Pairing_sex)) +
  geom_boxplot( alpha = 0.7) +
  scale_fill_manual(values = c("#F0CC4A","#a7e940", "#2EB89A", "#6a6542")) +
  labs(       x = "Pairing Sex",
    y = "Sniffing Duration (s)") +
  theme_classic() + theme(legend.position = "none")+ theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9))


#Individual variation
k <-ggplot(sniff, aes(y = Subject, x = Time_sec, fill = Subject)) +
    geom_boxplot(width = 0.5, alpha = 0.6) +
  scale_fill_manual(values = my_colors) +
  labs(       y = "Rhino ",
    x = "Sniff Duration (s)") +
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) + theme(legend.position = "none") + xlim(0, 17)
```

graph on sniffing duration depending on day of the experiment per pairing 

```{r}
# Mean and standard error per day per pairing type
sniff_summary <- sniff %>%
  group_by(Pairing_Day, Subject) %>%
  summarise(
    mean_time = mean(Time_sec),
    se_time = sd(Time_sec)/sqrt(n()),
    .groups = "drop")

# plot
ggplot(sniff_summary, aes(x = Pairing_Day, y = mean_time, color = Subject, group = Subject)) +
  geom_line(size = 0.7) +
  geom_point(size = 1.7) +
  labs(    x = "Pairing Day",
    y = "Mean Sniff Duration (s)",
    color = "Subject") +
  theme_minimal(base_size = 14)


#Mean times 
ggplot(sniff, aes(x = Pairing_Day, y = Time_sec )) +
  geom_line(data = sniff_summary, aes(x= Pairing_Day, y= mean_time, group = Subject, color = Subject), size = 0.8, alpha = 0.3) + scale_color_manual(values = my_colors) +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "#4b1957", size = 1.2) +
  stat_summary(fun = mean, geom = "point",
               color = "#4b1957", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.1, color = "#4b1957") +
   annotate("text", x = c(1, 2, 3), y = c( 5.7, 4.7, 4.8 ), 
           label = c("a", "b", "b"),  color = "#4b1957", size = 5) +
  scale_x_discrete(expand = c(0.1, 0.1)) + 
  labs(x = "Presentation Day",
       y = "Mean Sniff Duration (s)") +
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) 


ggplot(sniff_summary, aes(x = Pairing_Day, y = mean_time, color = Subject, group = Subject)) +
  geom_line(size = 0.8, alpha = 0.7) +
  geom_ribbon(aes(ymin = mean_time - se_time, ymax = mean_time + se_time, fill = Subject),
              alpha = 0.1, color = NA) +
  scale_color_manual(values = my_colors) +
  labs(    x = "Presentation Day",    y = "Mean Sniff Duration (s)",
    color = "Focal Rhino",    fill = "Focal Rhino") +
theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) 
```

#Dung scraping

```{r}

scrap <- beh %>%
  filter(Code == "Scrap") %>%
  group_by( Subject, Pairing_sex, Pairing_Day, Subject_sex, Dungfrom) %>%   
  summarise(Scrap_Count = n(), .groups = "drop")

hist(scrap$Scrap_Count)

fit <- glm(Scrap_Count ~ Dungfrom,
            family = poisson(),
            data = scrap)
Anova(fit)# no significative
anova(fit)

fit <- glm(Scrap_Count ~ Pairing_sex,
            family = poisson,
            data = scrap)
Anova(fit)# no significative


```


#Pooping on poop

```{r}

poop <- beh %>%
  filter(Code == "Poop") %>%
  group_by( Subject, Pairing_sex, Pairing_Day, Subject_sex, Dungfrom) %>%   
  summarise(Poop_Count = n(), .groups = "drop")

fit <- glmer(  Poop_Count ~ Dungfrom + (1|Subject),
  family = poisson, data = poop)
Anova(fit) #not significant

#check if significant difference 
fit <- glmer(  Poop_Count ~ Pairing_sex + Pairing_Day + (1|Subject),
  family = poisson, data = poop)
Anova(fit)# no significative


```

Graphs 

```{r}

c <- ggplot(poop, aes(x = Subject_sex, y = Poop_Count, fill = Subject_sex)) +
    geom_boxplot(alpha = 0.7) +
  labs(      x = "Rhino Sex",
    y = "Countermarking Freq." ) +
  scale_fill_manual(values = c("#a51c1c", "#f7ab47")) +
  theme_classic() + theme(legend.position = "none") + theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9))


d <- ggplot(scrap, aes(x = Subject_sex, y = Scrap_Count, fill = Subject_sex)) +
    geom_boxplot(alpha = 0.7) +
  labs(      x = "Rhino Sex",
    y = "Scrapping Frequency" ) +
  scale_fill_manual(values = c("#a51c1c", "#f7ab47")) +
  theme_classic() + theme(legend.position = "none") + theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9))


x <- ggplot(scrap, aes(x = Pairing_sex, y = Scrap_Count, fill = Pairing_sex)) +
    geom_boxplot(alpha = 0.7) +
  labs(    x = "Pairing Sex",
    y = "Scrapping Frequency" ) +
  scale_fill_manual(values = c("#F0CC4A","#a7e940", "#2EB89A", "#6a6542")) +
  theme_classic()  + theme(legend.position = "none") + theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9))


w <- ggplot(poop, aes(x = Pairing_sex, y = Poop_Count, fill = Pairing_sex)) +
    geom_boxplot(alpha = 0.7) +
  labs(    x = "Pairing Sex",
    y = "Countermarking Freq." ) +
  scale_fill_manual(values = c("#F0CC4A","#a7e940", "#2EB89A", "#6a6542")) +
  theme_classic()  + theme(legend.position = "none") + theme(axis.text.y = element_text(size = 9)) + theme(axis.title.x = element_text(size = 10))  + theme(axis.title.y = element_text(size = 9)) + theme(axis.title.y = element_text(size = 9))
```
Okay now the merged graphs :) 

```{r}
# Subject sex 

boxplotdryvol<-ggarrange(b, a, c, d,    labels = c("A", "B", "C", "D"), ncol = 2, nrow = 2, common.legend = F)

ggarrange(b, a, c,d,  labels = c("A", "B", "C", "D"), ncol = 2, nrow = 2, common.legend = F)

# Dung sex 


# Sex pairing 
boxplotdryvol<-ggarrange( y,z, w,x,    labels = c("A", "B", "C", "D"), ncol = 2, nrow = 2, common.legend = F)

ggarrange( y, z,  w, x,  labels = c("A", "B", "C", "D"), ncol = 2, nrow = 2, common.legend = F)

# Individual variation

boxplotdryvol<-ggarrange(j,k,    labels = c("A", "B"), ncol = 2, nrow = 2, common.legend = F)

ggarrange(j,k,  labels = c("A", "B"), ncol = 2, nrow = 2, common.legend = F)

```

