---
title: "Ona_voc_code"
author: "Ona Corberó Forn"
date: "2026-05-28"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(tidyverse)
library(plotly)
library(tibble)
library(viridis)
library(ggridges)
library(forcats) 
library(randomForest)
library(caTools)
library(rfPermute)
library(caret)
library(ggsignif)
library(ggpubr)
```

## VOCs Analysis from the Black Rhino Dung

Charge the necessary packages and set up the common used paths to the working directory 

```{r}

wd <- list()
wd$data   <- "D:/ona/biology/Master Biologie/2nd year/Rhino Project/Corbero_Ona_thesis/datasets/"
wd$output <- "D:/ona/biology/Master Biologie/2nd year/Rhino Project/Corbero_Ona_thesis/"

```

Filter the petrol molecules and calibrate the VOCs dataset with the IS: 

```{r}
vocs <- read.csv(paste0(wd$data, "ona_vocs.csv"))

vocs$row.retention.time <- round(vocs$row.retention.time , digits = 3) # retention time row with  3 digits
vocs2 <- vocs %>% select(-c(1,2))   %>%  column_to_rownames(var = "row.retention.time") %>%  t #remove first two columns and transpose

rownames(vocs2)<- rownames(vocs2) %>% str_replace(".CDF.Peak.area*", "") #replace .Peak.area

###fill missing values for IS
which(vocs2[,60] ==0) 
vocs2 <- as.data.frame(vocs2)
names(vocs2)[60] <- c("IS")

vocs3 <- vocs2%>% mutate(IS = ifelse(IS == 0, mean(IS[IS != 0]), IS)) #callibrate with IS 
vocs3$IS

vocs4 <- vocs3/vocs3$IS * 2
vocs5 <- vocs4[,-60] #new database without naphtalene 

rownames(vocs5) # Identify the columns where 'Dichloro.CDF Peak area' row has a value greater than 0
vocs5["imp", ] <- vocs5["Dichloro", ] + vocs5["DichloNa", ] + vocs5["Test", ]

columns_to_keep <- which(vocs5["imp", ] == 0)

vocs5_filtered <- vocs5[, columns_to_keep] # Subset the data frame to keep only those columns

vocs5_filtered2 <- vocs5_filtered %>% filter(!(rownames(vocs5_filtered) %in% c("imp", "Dichloro", "DichloNa", "Test")))

```

Calibrate with sample dry weight (dw) and merge with the dataset: 

```{r}
metadata<-read.csv(paste0(wd$data,  "ona_metadata_vocs.csv"), h=T,  sep=";",  na.strings = "NA", stringsAsFactors=FALSE)

metadata$rhino <- as.factor(metadata$rhino)
metadata$sample_type <- as.factor(metadata$sample_type)
metadata$rhino_sex <- as.factor(metadata$rhino_sex)
metadata$rhino_origin <- as.factor(metadata$rhino_origin)
metadata$day_exp <- as.factor(metadata$day_exp)
metadata$dung_rhino <- as.factor(metadata$dung_rhino)
metadata$dung_sex <- as.factor(metadata$dung_sex)
metadata$dung_origin <- as.factor(metadata$dung_origin)
metadata$date2 <- as.Date(metadata$date2, format = "%d/%m/%Y")

metadata2 <- metadata %>%  column_to_rownames(var = "filename")

#callibrating dw 

dw_values <- metadata2[rownames(vocs5_filtered2), "dw"]
dw_values <- as.numeric(dw_values)
vocs5_calibrated <- vocs5_filtered2/dw_values*100

# adding names to the compounds !!!!

compound_names<-read.csv(paste0(wd$data,  "names_voc_compounds.csv"), h=T,  sep=";",  na.strings = "NA", stringsAsFactors=FALSE)

vocs_names <- vocs5_calibrated %>%
  tibble::rownames_to_column("Sample") %>%
  pivot_longer(-Sample,   names_to = "Retention_time",
    values_to = "Composition") %>%
  mutate(Retention_time = as.numeric(Retention_time)) %>%
  left_join(compound_names, by = "Retention_time") %>%
 
  filter(!is.na(Compound_Name)) %>%  # Keep only retention times with a compound name
  
  # sum compositions for the same compound (per sample)
  group_by(Sample, Compound_Name) %>%
  summarise( Composition = sum(Composition, na.rm = TRUE), .groups = "drop") %>%
  
  # Make compound names columns
  pivot_wider(names_from = Compound_Name, values_from = Composition,
    values_fill = 0) %>%
  column_to_rownames("Sample")

##merging
tab_vocs <- merge(metadata2, vocs_names, by =0) 

##remove plant samples 

tab_vocs <- tab_vocs %>% filter(sample_type != "Plants")

#output_path <- file.path(wd$output, "ona_merged.csv")

#write.csv(tab_vocs, file = output_path, row.names = TRUE)

```

Data Analysis: 

# Non Metrical Dimensional Scaling (NDMS) 

To see the ordination of the samples spatially depending of similarity 

```{r}
names(tab_vocs)
mynmds <- metaMDS(tab_vocs[,c(20:56)], trace = FALSE,  k = 3, trymax = 100)
plot(mynmds)

### Extract NMDS site scores
nmds_sites <- as.data.frame(scores(mynmds, display = "sites"))

###Add metadata
nmds_sites$Rhino <- tab_vocs$rhino
nmds_sites$ID <-tab_vocs$Row.names
nmds_sites$Sex <-tab_vocs$rhino_sex
nmds_sites$Age <-tab_vocs$rhino_age
nmds_sites$Age <-factor(nmds_sites$Age, levels = c("4","5","6","8"))
nmds_sites$Rhino <- factor(nmds_sites$Rhino, levels = c( "Billie", "FatAmy", "Joko", "Mafukuzela",
                             "Mpilo", "Mshanelo", "Ndlebe", "Nsele","Qaphela", "Somiso", "TheRock"))

#### Extract species scores
nmds_species <- as.data.frame(scores(mynmds, display = "species"))

ggplot(nmds_sites, aes(x = NMDS1, y = NMDS2, label = Rhino)) +
 geom_point(size = 3, color = "blue") +
 geom_text(vjust = -1, size = 5) +
  geom_text(data = nmds_sites, aes(x = NMDS1, y = NMDS2, label = ID), color = "red", size = 4) +
  theme_minimal() +
  labs(x = "NMDS Axis 1", y = "NMDS Axis 2") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#### check where are the outliers
which(tab_vocs$Row.names =="ThMs3")
which(tab_vocs$Row.names =="MpTh17")
which(tab_vocs$Row.names =="Jo7")

##Create palette
my_colors1 <- c("#a51c1c", "#f7ab47", "#fff743", "#00f570", "#0a9341", "#a7ffe2", "#1de8f5", "#0151ca", "#5f01ca", "#f03ad3", "#f51d64")

my_colors <- c( "#a51c1c", "#ffe911", "#a845d3", "#0151cb","#F18F01", "#1a9cd7","#7F5539" ,"#00f580", "#F15BB5", "#1B334B", "#6A994E")

#NMDS sorted by Rhino name (without outliers) 
ggplot(nmds_sites [-c(297, 138),], aes(x = NMDS1, y = NMDS2, color = Rhino)) +
  geom_point(size = 2, alpha = 0.7) +  
  stat_ellipse(geom = "polygon",
               aes(fill = Rhino), level = 0.7, 
               alpha = 0.15, color = NA)+
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) + 
  theme_classic() +  labs(x = "NMDS Axis 1", y = "NMDS Axis 2", fill = "Rhino ID", color = "Rhino ID") +
  theme(plot.title = element_text(hjust = 0.5))


# 3D plot (to visuaise in 3d space)
plot_ly(nmds_sites[-c(297, 138),], 
        x = ~NMDS1, y = ~NMDS2, z = ~NMDS3,  
        color = ~Rhino,  colors = my_colors,  
        type = "scatter3d", mode = "markers", 
        marker = list(size = 5, opacity = 0.7)) %>%
  layout(title = "3D NMDS Plot for VOCs Black Rhino Dung",
         scene = list(xaxis = list(title = "NMDS Axis 1"),
           yaxis = list(title = "NMDS Axis 2"), zaxis = list(title = "NMDS Axis 3")))


```
# Plot by sex 

```{r}
ggplot(nmds_sites [-c(297, 138),], aes(x = NMDS1, y = NMDS2, color = Sex)) +
  geom_point(size = 2, alpha = 0.6) +  
  stat_ellipse(geom = "polygon",  aes(fill = Sex),  alpha = 0.2, color = NA, level = 0.95)+
  scale_color_manual(values = my_colors1) +
  scale_fill_manual(values = my_colors1) + 
  labs( x = "NMDS Axis 1", y = "NMDS Axis 2", fill = "Rhino Sex", color = "Rhino Sex") +
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  +  theme(axis.text = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) 


```
#Plot by age

```{r}

ggplot(nmds_sites [-c(297, 138),], aes(x = NMDS1, y = NMDS2, color = Age)) +
  geom_point(size = 2, alpha = 0.8) +  
  stat_ellipse(geom = "polygon",
               aes(fill = Age), 
               alpha = 0.15, color = NA)+
  scale_color_manual(values = c ("#75E0EB", "#4C55AE", "#BB35C5", "#F96783")) +
  scale_fill_manual(values = c ("#75E0EB", "#4C55AE","#BB35C5","#F96783" )) + 
  labs( x = "NMDS Axis 1", y = "NMDS Axis 2", fill = "Rhino Age", color = "Rhino Age") +
   theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  +  theme(axis.text = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11))


```

# Most abundant VOCs 

To know the distribution of the volatile molecules found in the samples


```{r}

#volatile abundance across all samples

volatiles <- colSums(tab_vocs[, c(20:56)], na.rm = TRUE)

compound_function<-read.csv(paste0(wd$data,  "voc_functionalgroups.csv"), h=T,  sep=";",  na.strings = "NA", stringsAsFactors=FALSE)

# to set as dataframe
volatile_df <- data.frame( compound_names = names(volatiles),  Abundance = volatiles)

volatile_info <- merge(volatile_df, compound_function, by = "compound_names", all.x= TRUE)


# Abundance of compounds colored by functional group 
ggplot(volatile_info[1:37,], aes(x = reorder(compound_names, -Abundance), y = Abundance, fill = Functional_group)) +  geom_bar(stat = "identity") +
  labs( x = "VOCs", y = "Total Abundance" ) +
  theme_classic() + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# grouped by functional group 
abundance_functional <- volatile_info %>%
  group_by(Functional_group) %>%
  summarise(Total_Abundance = sum(Abundance, na.rm = TRUE))

ggplot(abundance_functional, aes(x = Total_Abundance, y = reorder(Functional_group, Total_Abundance), fill = Functional_group)) +
  geom_bar(stat = "identity") + labs( x = "Total Abundance",
    y = " Functional Groups" ) +  scale_fill_brewer(palette = "Set3") +
  theme_classic() +   theme(axis.title.y = element_blank()) + theme (legend.position = "none")


```
# Variability of each rhino profile 

To visualise the variance for each rhinoceros chemical profile 

```{r}
##### with all samples for one rhino

billie_voc_long <- tab_vocs %>%
  filter(rhino == "Billie") %>%  
  select(Row.names, 20:56) %>%  
  pivot_longer(cols = -Row.names, names_to = "VOCs", values_to = "Abundance")

billie_voc_long$Row.names <- as.factor(billie_voc_long$Row.names)

## Plot with separate plots per sample
ggplot(billie_voc_long, aes(x = as.factor(VOCs), y = Abundance, fill = Row.names)) + geom_bar(stat = "identity") +
  facet_wrap(~ Row.names, scales = "free_y") +  
  labs( x = "VOCs", y = "Abundance") +
  theme_minimal() + theme(    axis.text.x = element_blank(), 
    legend.position = "none") + scale_fill_viridis(discrete=TRUE, option="A")

#Plot for one sample 
billie_one_sample <- tab_vocs %>%
  filter(rhino == "Billie", Row.names == "Bi9") %>%
  select(Row.names, 20:56) %>%
  pivot_longer(cols = -Row.names, names_to = "compound_names", values_to = "Abundance")

billie_merged <- merge(  billie_one_sample,  volatile_info[, c("compound_names", "Retention_time")],
  by = "compound_names",  all.x = TRUE)

ggplot(billie_merged, aes(x = Retention_time, y = Abundance)) +
  geom_segment(aes(xend = Retention_time, yend = 0), 
    linewidth = 0.3, color = "blue") +
  labs(x = "Retention time (min)", y = "Abundance") +  theme_classic()
```

#  PERMANOVA analysis!!!! 

To assess which variables (identity, sex, age, etc..) explain the most the variation in VOCs composition among the samples!

```{r}
fit<-adonis2(tab_vocs[,c(20:56)]~rhino, data= tab_vocs, permutations = 999,method = "bray") 
fit # Rhino ID very significant effect on all sample types !

## now let's verify with permdist 

dist_mat <- vegdist(tab_vocs[, 20:56], method = "bray")

disp <- betadisper(dist_mat, tab_vocs$rhino)

disp_test <- permutest(disp, permutations = 999)

disp
disp_test # not significant, so no differences in dispersion among rhino groups :) 

```
```{r}

fit<-adonis2(tab_vocs[,c(20:56)]~rhino_sex, data= tab_vocs, permutations = 999,method = "bray") 
fit # Rhino sex significant!! but not as much as ID

## now let's verify with permdist 

dist_mat <- vegdist(tab_vocs[, 20:56], method = "bray")

disp <- betadisper(dist_mat, tab_vocs$rhino_sex)

disp_test <- permutest(disp, permutations = 999)

disp
disp_test # not significant, so no differences in dispersion among rhino sex :) 

```
```{r}

fit<-adonis2(tab_vocs[,c(20:56)]~rhino_age, data= tab_vocs, permutations = 999,method = "bray") 
fit # also significant but small effect

## now let's verify with permdist 

dist_mat <- vegdist(tab_vocs[, 20:56], method = "bray")

disp <- betadisper(dist_mat, tab_vocs$rhino_age)

disp_test <- permutest(disp, permutations = 999)


disp
disp_test #  not significant!!

```
```{r}
fit<-adonis2(tab_vocs[,c(20:56)]~sample_type, data= tab_vocs, permutations = 999,method = "bray") 
fit # Rhino ID has a verY significant effect on VOC composition! and sample type, and the two combined vary between the different rhinoceros 



```

# PERMANOVA for each rhino separately to see the intersection effect of ID and the sample type 

Using the Bray-Curtis dissimilarity, to see how sample type (so experiment vs baseline) affects variance in the VOCs composition in the samples from each rhino. So if sample type has an effect on the variance of their chemical profiles. 

```{r}
unique_rhinos <- unique(tab_vocs$rhino) # to create list of rhinos

# permanova for each rhino separately
results2 <- lapply(unique_rhinos, function(rhino_id) {
  subset_data <- tab_vocs[tab_vocs$rhino == rhino_id, ]  
  
  if (nrow(subset_data) > 1) {  
    fit <- adonis2(subset_data[, c(20:56)] ~ sample_type, 
  data = subset_data, permutations = 999, method = "bray")
    return(list(rhino = rhino_id, fit = fit))
  } else {
    return(list(rhino = rhino_id, fit = NULL))  
  }
})

# Results per rhino
for (res in results2) {
  if (!is.null(res$fit)) {
    print(paste("Rhino:", res$rhino))
    print(res$fit)
  } else {
    print(paste("Rhino:", res$rhino, "- Not enough data for analysis"))
  }
}
```

So we can see that for all rhinos sample type has a significant effect on VOC variation, explaining between the 20-40% of the variance. So, sample type has a significant effect on VOC variance. 


# Dispersion and PCoA

To know which sample type has more variance 

```{r}
tab_vocs_filtered <- tab_vocs[tab_vocs$sample_type %in% c("Baseline", "Experiment"), ]

bray_dist <- vegdist(tab_vocs_filtered[, 20:56], method = "bray") # dissimilarity between samples depending on VOC abundance 

dispersion <- betadisper(bray_dist, tab_vocs_filtered$sample_type) # mean dispersion (variance) for sample_type 

permutest(dispersion, permutations = 999) # to see if difference among sample type variances is significant -> yes it is!!!!! 


# Plot! 
pcoa_res <- cmdscale(bray_dist, eig = TRUE, k = 2)  

# data frame 
pcoa_df <- data.frame(
  PC1 = pcoa_res$points[,1],
  PC2 = pcoa_res$points[,2],
  Sample_Type = tab_vocs_filtered$sample_type)

#  % variance explained
var_explained <- round(100 * pcoa_res$eig / sum(pcoa_res$eig), 1)

#  PCoA
ggplot(pcoa_df, aes(x = PC1, y = PC2, color = Sample_Type, fill = Sample_Type)) +
  geom_point(shape = 21, size = 2, alpha = 0.7) +
  stat_ellipse(aes(group = Sample_Type), type = "t", alpha = 0.2, geom = "polygon", level = 0.95, color = NA) +
  labs(    x = paste0("PCoA1 (", var_explained[1], "%)"),
    y = paste0("PCoA2 (", var_explained[2], "%)") , fill = "Sample Type", color = "Sample Type" ) +
  scale_color_manual(values = c("#4C78A8", "gold")) +
  scale_fill_manual(values = c( "#4C78A8", "gold")) +
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  +  theme(axis.text = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) 

#Boxplot graph 
disp_df <- data.frame(  Sample_Type = tab_vocs_filtered$sample_type,
  Dispersion = dispersion$distances)

ggplot(disp_df, aes(x = Sample_Type, y = Dispersion, fill = Sample_Type)) +
  geom_boxplot(alpha = 0.5, color = "grey40", outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.6, aes(color = Sample_Type)) +  
  labs(    x = NULL,    y = "Variance (Dispersion)") +
  scale_fill_manual(values = c("#4C78A8", "gold")) +  
  scale_color_manual(values = c("#4C78A8", "gold")) + 
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  +  theme(axis.text = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) + theme(legend.position = "none") 


```

# ALPHA DIVERSITY within rhino between experiment and baseline (sample types) 

So we can see if the experiment affects on the diversity of VOCs that can be found in the dung of each rhinoceros 


```{r}
tab_vocs$shannon_diversity <- diversity(tab_vocs[, 20:56], index = "shannon") # shannon diversity index
tab_vocs_filtered$shannon_diversity <- diversity(tab_vocs_filtered[, 20:56], index = "shannon") # shannon diversity index

#wilcoxon paired tests with shannon, richness and evenness

tab_vocs_filtered <- tab_vocs_filtered %>%
  mutate(    richness = specnumber(tab_vocs_filtered[, 20:56]),
    evenness = shannon_diversity / log(richness))

avg_metrics <- tab_vocs_filtered %>% group_by(rhino, sample_type) %>%
  summarise(    mean_richness = mean(richness),
    mean_evenness = mean(evenness),
    mean_shannon = mean(shannon_diversity),.groups = "drop")

wide_richness <- avg_metrics %>%
  select(rhino, sample_type, mean_richness) %>%
  pivot_wider(names_from = sample_type, values_from = mean_richness)

wide_evenness <- avg_metrics %>%
  select(rhino, sample_type, mean_evenness) %>%
  pivot_wider(names_from = sample_type, values_from = mean_evenness)

wide_shannon <- avg_metrics %>%
  select(rhino, sample_type, mean_shannon) %>%
  pivot_wider(names_from = sample_type, values_from = mean_shannon)

wilcox_richness <- wilcox.test(wide_richness$Baseline, wide_richness$Experiment, paired = TRUE)
wilcox_evenness <- wilcox.test(wide_evenness$Baseline, wide_evenness$Experiment, paired = TRUE)
wilcox_shannon <- wilcox.test(wide_shannon$Baseline, wide_shannon$Experiment, paired = TRUE)

wilcox_richness #Significant decrease in the number of VOCs during the experiment
wilcox_evenness #non Significant decrease in how evenly VOCs are distributed -> some compounds dominate more during the experiment
wilcox_shannon # Significant decrease overall in alpha diversity, combining richness & evenness.


```

# Plot Shannon Diversity Index per day per Rhinoceros 

To see diversity over time and also visualize the effect of the experiment with the separated individuals


```{r}
sampled_days <- tab_vocs %>%
  filter(!is.na(shannon_diversity))

#Shannon Diversity index over time for each rhinoceros

tab_vocs$rhino_ordered <- fct_reorder(tab_vocs$rhino, tab_vocs$shannon_diversity)

ggplot(tab_vocs, aes(  x = date2,  y = rhino_ordered,   height = shannon_diversity / 2,
  group = rhino_ordered,   scale = 4,   fill = rhino_ordered)) +
  geom_ridgeline(    alpha = 0.3, scale = 2.5,  color = "white") +
  xlim(as.Date("2024-07-12"), as.Date("2024-08-25")) +  # Proper date format
  labs(  y = "Shannon diversity index", x = "Date", fill = "Rhino ID"  ) +
  geom_vline(xintercept = as.Date("2024-07-31"), color = "white", size = 0.5) +
  geom_vline(xintercept = as.Date("2024-08-17"), color = "white", size = 0.5) +
    scale_fill_manual(values = c("#a51c1c", "#F15BB5",  "#6A994E", "#a845d3", "#7F5539", "#1a9cd7","#0151cb","#1B334B",  "#F18F01", "#ffe911","#00f580")) +
   theme_classic() +  theme(axis.text.y = element_blank()) + theme(axis.title.x = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) + theme(axis.ticks.y = element_blank())



```
#compute variance coefficient 

```{r}
cv_by_rhino <- tab_vocs_filtered %>%
  group_by(rhino, sample_type ) %>%   
  summarise(    mean_shannon = mean(shannon_diversity, na.rm = TRUE),
    sd_shannon   = sd(shannon_diversity, na.rm = TRUE),
    cv_shannon   = sd_shannon / mean_shannon,
    n = n(), .groups = "drop")

cv_by_rhino #table 

#plot 
ggplot(cv_by_rhino,  aes(x = sample_type, y = cv_shannon, group = rhino, color = rhino)) +
  geom_point(size = 2.4, alpha = 0.7) +  geom_line(size= 0.4, alpha = 0.5) +
  scale_color_manual(values = my_colors) +
  labs(x = "Sample Type",
       y = "Coefficient of Variation (Shannon)", color = "Rhino ID" ) +
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11))


#to make new graph 

cv_change <- cv_by_rhino %>%
  select(rhino, sample_type, cv_shannon) %>%
  pivot_wider(names_from = sample_type, values_from = cv_shannon) %>%
  mutate(delta_cv = Experiment - Baseline)


ggplot(cv_change, aes(y = rhino, x = delta_cv, fill = rhino)) +
  geom_col(width = 0.6, alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c("#a51c1c", "#ffe911", "#a845d3", "#0151cb","#F18F01", "#1a9cd7","#7F5539" ,"#00f580", "#F15BB5", "#1B334B", "#6A994E")) +
  labs(    y = NULL,    x = "Δ CV (Experiment − Baseline)",
    fill = "Rhino ID") +
  theme_classic() +
  theme(axis.text.y.right  = element_text(size = 9 ), axis.ticks.y = element_blank(), axis.title.x = element_text(size = 11 ), axis.title.y = element_text(size = 11 ), legend.position = "none")



```

#Variance within the experiment 

Check which variables are the ones affecting on the VOC variance during the experiment (rhino sex, pairing sex, dung sex )

```{r}
experiment_data <- tab_vocs[tab_vocs$sample_type == "Experiment", ]
experiment_data <- experiment_data[experiment_data$rhino_sex != "", ]
experiment_data$rhino_sex <- droplevels(experiment_data$rhino_sex)

#output_path <- file.path(wd$output, "experiment_data.csv")
#write.csv(experiment_data, file = output_path, row.names = TRUE)

experiment_data2 <- read.csv(paste0(wd$data, "experiment_withpairingday.csv"), h=T,  sep=";",  na.strings = "NA", stringsAsFactors=FALSE)

experiment_data$pairing_day <- experiment_data2$pairing_day[
  match(experiment_data$short_id, experiment_data2$short_id)]

#permanova 

fit <- adonis2(experiment_data[, c(20:56)] ~ pairing_sex, data = experiment_data, 
               permutations = 999,  method = "bray", strata = experiment_data$rhino)
fit # pairing sex not significant once its restricted to include rhinos identity

fit <- adonis2(experiment_data[, c(20:56)] ~ day_exp, data= experiment_data, 
               permutations = 999,  method = "bray", strata = experiment_data$rhino)
fit # significant! should check pairing day 

fit <- adonis2(experiment_data[, c(20:56)] ~ pairing_day*dung_rhino, data= experiment_data, 
               permutations = 999,  method = "bray", strata = experiment_data$rhino , by ="terms")
fit # not significant

fit <- adonis2(experiment_data[, c(20:56)] ~ pairing_id, data = experiment_data, 
               permutations = 999,  method = "bray")
fit #significant!!! very important actually bc it means each rhino has a significantly different profile depending of who it's smelling, more than sex or other things ! biig effect also


fit <- adonis2(experiment_data[, c(20:56)] ~ rhino, data = experiment_data, 
               permutations = 999,  method = "bray", strata = experiment_data$rhino_sex)
fit # quite significant! -z rhinos have different profiles independent of sex


adonis_summary <- as.data.frame(fit)%>% rownames_to_column("Factor") %>%
  rename(     Df = Df,     SumOfSqs = SumOfSqs,
    R2 = R2, Fvalue = F, Pvalue = `Pr(>F)` ) %>%
  mutate(Significance = case_when(Pvalue <= 0.001 ~ "***",  Pvalue <= 0.01  ~ "**", Pvalue <= 0.05  ~ "*",
    Pvalue <= 0.1   ~ ".", TRUE            ~ ""))

adonis_summary
```
Can I visualise it? 

```{r}
mynmds2 <- metaMDS(experiment_data[,c(20:56)], trace = FALSE,  k = 3, trymax = 100)
plot(mynmds2)

# Extract NMDS site scores
nmds_sites2 <- as.data.frame(scores(mynmds2, display = "sites"))

#Add metadata
nmds_sites2$pairing_sex <- experiment_data$pairing_sex
nmds_sites2$pairing_id <-experiment_data$pairing_id
nmds_sites2$short_id <-experiment_data$short_id
#NMDS of the sex pairing 

ggplot(nmds_sites2, aes(x = NMDS1, y = NMDS2, label = pairing_sex)) +
 geom_point(size = 3, color = "blue") +
 geom_text(vjust = -1, size = 5) +
  geom_text(data = nmds_sites2, aes(x = NMDS1, y = NMDS2, label = short_id), color = "red", size = 4) +
  theme_minimal() +
  labs( x = "NMDS Axis 1", y = "NMDS Axis 2") +
  theme(plot.title = element_text(hjust = 0.5))

which(experiment_data$Row.names =="ThMs3")
which(experiment_data$Row.names =="MpTh17")

ggplot(nmds_sites2[-c(189, 88),], aes(x = NMDS1, y = NMDS2, color = pairing_sex)) +
  geom_point(size = 2, alpha = 0.8) +  # Plot points
  stat_ellipse(geom = "polygon",
               aes(fill = pairing_sex), 
               alpha = 0.15, color = NA ) +
  scale_color_manual(values = c("#F0CC4A","#a7e940", "#2EB89A", "#6a6542")) +
  scale_fill_manual(values = c("#F0CC4A","#a7e940", "#2EB89A", "#6a6542")) + 
   labs(x = "NMDS Axis 1", y = "NMDS Axis 2", fill = "Sex Pairing", color = "Sex Pairing") +
  theme(plot.title = element_text(hjust = 0.5)) + theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 11))  +  theme(axis.text = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) 


```



```{r}
mexp_data <- experiment_data[experiment_data$rhino_sex == "m", ]
fexp_data <- experiment_data[experiment_data$rhino_sex == "f", ]

# compute distances 
dist_m <- vegdist(mexp_data[, 20:56], method = "bray")
dist_f <- vegdist(fexp_data[, 20:56], method = "bray")

#calculate dispersion (permdisp) by pairing sex

disp_m <- betadisper(dist_m, mexp_data$pairing_sex)
disp_f <- betadisper(dist_f, fexp_data$pairing_sex)


# Overall test for differences in dispersion
anova(disp_m) #significative for males, so one of the pairing_sex causes different variability in VOCs
anova(disp_f) # not significative for females

# Post-hoc pairwise comparisons 
permutest(disp_m, pairwise = TRUE, permutations = 999) #male rhinos more variable when exposed to female dung


```

# visualisation of voc variation depending of sex pairing


```{r}

m_beta <- data.frame(  distance_to_centroid = disp_m$distances,
  pairing_sex = mexp_data$pairing_sex, rhino = mexp_data$rhino)

f_beta <- data.frame(  distance_to_centroid = disp_f$distances,
  pairing_sex = fexp_data$pairing_sex, rhino = fexp_data$rhino)


#Graph of Beta Diversity (Variability) of VOCs in Males
ggplot(m_beta, aes(x = pairing_sex, y = distance_to_centroid, fill = rhino)) +
  geom_boxplot(alpha = 0.5, color = "grey30", outlier.shape = NA)  +
  scale_fill_manual(values = c("#F18F01", "#1a9cd7","#00f580", "#1B334B", "#6A994E")) +
  labs(       x = "Sex Pairing",
       y = "Distance to Group Centroid (Bray-Curtis)", fill = "Male Rhinos") +  
   theme ( panel.grid.major.x = element_blank(), panel.grid.minor.y = element_blank() ) + 
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 0))  +  theme(axis.text = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) 

#Graph of Beta diversity (variabiliy) of VOCs in females 

ggplot(f_beta, aes(x = pairing_sex, y = distance_to_centroid, fill = rhino)) +
  geom_boxplot(alpha = 0.5, color = "grey30", outlier.shape = NA) +
  scale_fill_manual(values = c ("#a51c1c", "#ffe911", "#a845d3", "#0151cb", "#7F5539", "#F15BB5")) +  
  labs(    x = "Sex Pairing",
    y = "Distance to Group Centroid (Bray–Curtis)", fill = "Female Rhinos") +   
 theme ( panel.grid.major.x = element_blank(), panel.grid.minor.y = element_blank() ) + 
  theme_classic() +  theme(axis.text.y = element_text(size = 10)) + theme(axis.title.x = element_text(size = 0))  +  theme(axis.text = element_text(size = 11))  + theme(axis.title.y = element_text(size = 11)) 


```

####  RandomForest 

To establish links between specific volatiles and individual characteristics, and check which volatiles can be predictors of them 

```{r}

baseline_data <- tab_vocs %>%
  filter(sample_type == "Baseline") %>%
  mutate(rhino_sex = droplevels(rhino_sex))

# rhino_sex and quantification columns
rf_data <- baseline_data %>%
dplyr::select(rhino_sex, 20:56) %>%
  na.omit()

names(rf_data) <- make.names(names(rf_data), unique = TRUE)

# Fit Random Forest to the train data set
set.seed(123)  # Setting seed
classifier_RF = randomForest(rhino_sex ~ .,
  data = rf_data,
  ntree = 1000,
  importance = TRUE)

print (classifier_RF)

# Plotting Error Rate
plot(classifier_RF, main = "Random Forest Error Rate") 

# Importance plot
imp<- importance(classifier_RF)
imp <- as.data.frame(imp)
imp <- tibble::rownames_to_column(imp, "Volatiles")

# Sort by MeanDecreaseGini
imp_sorted <- imp[order(imp$MeanDecreaseGini,decreasing=TRUE),]

# Variable importance plot
varImpPlot(classifier_RF)



```

# RF Permute
One more randomforest run to check the statistical significance of the model and the predictor variable
```{r}

set.seed(123)
rf_perm <- rfPermute(
  rhino_sex ~ .,
  data = rf_data,
  ntree = 500,
  num.rep = 100)

rf_perm # to check model significance 

rf_perm$pval #to see all the  p-values

rf_perm$rf$importance



```
#Cross Validated Random forest 

To know if there is too much noise from the VOCs and that is why the model is performing poorly, or to know if there are only no specific VOC that can help separate sexes we do a cross Validated Random Forest 



```{r}
#Cross Validated Random Forest  

ctrl <- trainControl(  method = "repeatedcv",  number = 10,  repeats = 5)

set.seed(123)

rf_cv <- train(  rhino_sex ~ .,  data = rf_data,  method = "rf",  trControl = ctrl,  tuneLength = 10)

rf_cv #to check model accuracy -> seems accurate for sex! 


###VOCs with significant permutation importance

# Extract p-values from rfPermute
pvals<- rf_perm$pval[, , "unscaled"]

# Select variables with p < 0.05 for either class
sig_vars <- rownames(pvals)[apply(pvals, 1, function(x) any(x < 0.05))]

sig_vars

#dataset with significant VOCs 
rf_sig <- rf_data[, c("rhino_sex", sig_vars)]

```

#Importance plot + significance plot 

most important compounds for discrimination (according to what class (sex/age) you selected above), its the same as the classifier plot above based on MeanDecreaseGini
```{r}
# top 10 VOCs
top10_vocs <- imp_sorted[1:10, ]
print(top10_vocs)

# Top 10 VOCs by importance (mean decrease accuracy) for sex 

 ggplot(top10_vocs, aes(x=reorder(Volatiles,MeanDecreaseAccuracy), y=MeanDecreaseAccuracy,fill=MeanDecreaseAccuracy))+ 
      geom_bar(stat="identity", position="dodge")+ coord_flip()+
      ylab("Mean Decrease Accuracy")+
      xlab("")+
      guides(fill=F)+
      scale_fill_gradient(low="grey", high="#27391C") +
      theme_classic() 
 
 
 #boxplots of top 5 significant volatiles: Male vs. Female

 # first check significance !
 
 rf_data$rhino <- rf_data2$rhino
 
fit <- glm(rf_data$X1.7.Dimethyl.4..1.methylethyl.cyclodecane ~ rhino_sex + rhino , family = quasipoisson, data = rf_data)
car::Anova(fit)
anova(fit) # significant for both rhino sex (**) & individuality (***)

fit <- glm(rf_data$Tridecane ~ rhino_sex + rhino , family = quasipoisson, data = rf_data)
car::Anova(fit)
anova(fit) #significant for rhino (***) but not for rhino sex 

fit <- glm(rf_data$Decane..2.6.6.trimethyl. ~ rhino_sex + rhino , family = quasipoisson, data = rf_data)
car::Anova(fit)
anova(fit)  #significant for rhino (***) but not for rhino sex 

fit <- glm(rf_data$X.6.Methyl.2.heptanone ~ rhino_sex + rhino , family = quasipoisson, data = rf_data)
car::Anova(fit)
anova(fit)# significant for both rhino sex (***) & individuality (**)

fit <- glm(rf_data$Dodecane ~ rhino_sex + rhino , family = quasipoisson, data = rf_data)
car::Anova(fit)
anova(fit) #significant for rhino (***) but not for rhino sex 

fit <- glm(rf_data$Nonadecane ~ rhino_sex + rhino , family = quasipoisson, data = rf_data)
car::Anova(fit)
anova(fit) # significant for both rhino sex (***) & individuality (**)

#plot

d<- ggplot(rf_data, aes(x=rhino_sex, y = `X.6.Methyl.2.heptanone`, fill = rhino_sex)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values=c("#a51c1c", "#f7ab47")) +
  labs(      y = "Relative abundance"  ) +
    theme_classic() + theme(axis.title.x = element_blank()) +  theme(legend.position = "none") +
  geom_signif(  annotations = "***",
  xmin = 1, xmax = 2,  y_position = max(rf_data$X.6.Methyl.2.heptanone)*0.95  ,  textsize = 3.5)

a<- ggplot(rf_data, aes(x=rhino_sex, y=	`X1.7.Dimethyl.4..1.methylethyl.cyclodecane`, fill = rhino_sex)) + 
  geom_boxplot(alpha = 0.7) + 
  scale_fill_manual(values=c("#a51c1c", "#f7ab47")) +
  labs(    y = "Relative abundance"  ) +
     theme_classic() + theme(axis.title.x = element_blank()) +  theme(legend.position = "none") +
  geom_signif(  annotations = "**",
  xmin = 1, xmax = 2,  y_position = max(rf_data$X1.7.Dimethyl.4..1.methylethyl.cyclodecane) *0.95 ,  textsize = 3.5)

e<- ggplot(rf_data, aes(x=rhino_sex, y=`Dodecane`, fill = rhino_sex)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values=c("#a51c1c", "#f7ab47")) +
  labs(    y = "Relative abundance"  ) +
   theme_classic() + theme(axis.title.x = element_blank()) +  theme(legend.position = "none") +
  geom_signif(  annotations = "ns",
  xmin = 1, xmax = 2,  y_position = max(rf_data$Dodecane) * 1.1,  textsize = 3.5)

f <- ggplot(rf_data, aes(x=rhino_sex, y=`Nonadecane`, fill = rhino_sex)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values=c("#a51c1c", "#f7ab47")) +
  labs(    y = "Relative abundance"  ) +
   theme_classic() + theme(axis.title.x = element_blank()) +  theme(legend.position = "none") +
  geom_signif(  annotations = "***",
  xmin = 1, xmax = 2,  y_position = max(rf_data$Nonadecane)*0.95 ,  textsize = 3.5)
  
b <- ggplot(rf_data, aes(x=rhino_sex, y=`Tridecane`, fill = rhino_sex)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values=c("#a51c1c", "#f7ab47")) +
  labs(     y = "Relative abundance"  ) +
  theme_classic() + theme(axis.title.x = element_blank()) +  theme(legend.position = "none") +
  geom_signif(  annotations = "ns",
  xmin = 1, xmax = 2,  y_position = max(rf_data$Tridecane) * 1.1,  textsize = 3.5)

c <- ggplot(rf_data, aes(x=rhino_sex, y=`Decane..2.6.6.trimethyl.`, fill = rhino_sex)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values=c("#a51c1c", "#f7ab47")) +
  labs(     y = "Relative abundance"  ) +
  theme_classic() + theme(axis.title.x = element_blank()) +  theme(legend.position = "none") +
  geom_signif(  annotations = "ns",
  xmin = 1, xmax = 2,  y_position = max(rf_data$Decane..2.6.6.trimethyl.) * 1.1,  textsize = 3.5)

boxplotdryvol<-ggarrange(a, b, c, d, e, f,   labels = c("A", "B", "C", "D", "E", "F"), ncol = 3, nrow = 2, common.legend = F)
ggarrange(a, b, c, d, e, f,   labels = c("A", "B", "C", "D", "E", "F"), ncol = 3, nrow = 2, common.legend = F)

 
```
And now for individuality 

```{r}

# Include rhino  and quantification columns
rf_data2 <- baseline_data %>%
dplyr::select(rhino, 20:56) %>%
  na.omit()
rf_data2$rhino <- droplevels(as.factor(rf_data2$rhino))
names(rf_data2) <- make.names(names(rf_data2), unique = TRUE)

### Fit Random Forest to the train data set
set.seed(123)  

classifier_RF = randomForest(rhino ~ .,
  data = rf_data2,
  ntree = 1000,
  importance = TRUE)

print (classifier_RF)

# Plotting Error Rate
plot(classifier_RF, main = "Random Forest Error Rate") 

# Importance plot
imp<- importance(classifier_RF)
imp <- as.data.frame(imp)
imp <- tibble::rownames_to_column(imp, "Volatiles")

# Sort by MeanDecreaseGini
imp_sorted <- imp[order(imp$MeanDecreaseGini,decreasing=TRUE),]

# Variable importance plot
varImpPlot(classifier_RF)

#### RF permute 

set.seed(123)
rf_perm <- rfPermute(
  rhino ~ .,
  data = rf_data2,
  ntree = 500,
  num.rep = 100)

rf_perm # to check model significance 

rf_perm$pval #extract p-values


 #### Cross validated model

ctrl <- trainControl(  method = "repeatedcv",  number = 10,  repeats = 5)

set.seed(123)

rf_cv <- train(  rhino ~ .,  data = rf_data2,  method = "rf",  trControl = ctrl,  tuneLength = 10)

rf_cv #to check model accuracy -> seems not super accurate 


# Significant VOCs

pvals <- rf_perm$pval[, , "unscaled"]

# Select variables with p < 0.05 for either class
sig_vars <- rownames(pvals)[apply(pvals, 1, function(x) any(x < 0.05))]

sig_vars

#dataset with only significant VOCs 
rf_sig2 <- rf_data2[, c("rhino", sig_vars)]

```

Importance plot for individuality 

```{r}
# View top 10 VOCs
top10_vocs2 <- imp_sorted[1:10, ]
print(top10_vocs2)

 ggplot(top10_vocs2, aes(x=reorder(Volatiles,MeanDecreaseAccuracy), y=MeanDecreaseAccuracy,fill=MeanDecreaseAccuracy))+ 
      geom_bar(stat="identity", position="dodge")+ coord_flip()+
      ylab("Mean Decrease Accuracy")+
      xlab("")+
      guides(fill=F)+
      scale_fill_gradient(low="grey", high="#27391C") +
      theme_classic() 

```
#And Age  

```{r}
rf_data3 <- baseline_data %>%
dplyr::select(rhino_age, 20:56) %>%
  na.omit()
rf_data3$rhino_age <- as.factor(rf_data3$rhino_age)
names(rf_data3) <- make.names(names(rf_data3), unique = TRUE)

### Fit Random Forest to the train data set
set.seed(123)  

classifier_RF = randomForest(rhino_age ~ .,
  data = rf_data3,
  ntree = 1000,
  importance = TRUE)

print (classifier_RF)

# Plotting Error Rate
plot(classifier_RF, main = "Random Forest Error Rate") 

# Importance plot
imp<- importance(classifier_RF)
imp <- as.data.frame(imp)
imp <- tibble::rownames_to_column(imp, "Volatiles")

# Sort by MeanDecreaseGini
imp_sorted <- imp[order(imp$MeanDecreaseGini,decreasing=TRUE),]

# Variable importance plot
varImpPlot(classifier_RF)

#### RF permute 

set.seed(123)
rf_perm <- rfPermute(
  rhino_age ~ .,
  data = rf_data3,
  ntree = 500,
  num.rep = 100)

rf_perm # to check model significance 

rf_perm$pval #extract p-values


 #### Cross validated model

ctrl <- trainControl(  method = "repeatedcv",  number = 10,  repeats = 5)

set.seed(123)

rf_cv <- train(  rhino_age ~ .,  data = rf_data3,  method = "rf",  trControl = ctrl,  tuneLength = 10)

rf_cv #to check model accuracy -> seems not super accurate 


# Significant VOCs

pvals <- rf_perm$pval[, , "unscaled"]

# Select variables with p < 0.05 for either class
sig_vars <- rownames(pvals)[apply(pvals, 1, function(x) any(x < 0.05))]

sig_vars

#dataset with only significant VOCs 
rf_sig3 <- rf_data3[, c("rhino_age", sig_vars)]

```

plot  

```{r}
# View top 10 VOCs
top10_vocs3 <- imp_sorted[1:10, ]
print(top10_vocs3)

 ggplot(top10_vocs3, aes(x=reorder(Volatiles,MeanDecreaseAccuracy), y=MeanDecreaseAccuracy,fill=MeanDecreaseAccuracy))+ 
      geom_bar(stat="identity", position="dodge")+ coord_flip()+
      ylab("Mean Decrease Accuracy")+
      xlab("")+
      guides(fill=F)+
      scale_fill_gradient(low="grey", high="#27391C") +
      theme_classic() 

```

